Rapid Epidemiological Analysis of Comorbidities and Treatments as risk factors for COVID-19 in Scotland (REACT-SCOT): a population-based case-control study

### 2020-06-01 casecontrol.Rmd
updated pharmacoepi analyses

_Background_-- Why some people infected with SARS-CoV-2 develop severe COVID-19 disease and others do not is poorly understood.  The objectives of this study were to identify risk factors for severe disease and to lay the basis for risk stratification based on demographic data and health records.

_Methods_ -- The design was a matched case-control study. Cases were ascertained as all those with a positive nucleic acid test for SARS-CoV-2 in the national virology laboratory database.  Those who had been admitted to a critical care unit or had died within 28 days of a positive test were classified as severe cases.  For each case, ten controls matched for sex, age and primary care practice were selected from the population register.  For cases and controls we extracted all diagnostic codes from the past five years of hospital discharge records, all drug codes from prescriptions encashed during the past nine months and we linked to the national diabetes register.  Rate ratios for associations with severe COVID-19 were estimated by conditional logistic regression.  The performance of predictive models for severe COVID-19 was estimated by cross-validation.

_Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.  In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]`.  Univariate rate ratios (95% CIs) for conditions listed by public health agencies as conferring high risk were  `r table.agegr.all[5, 3]` for Type 1 diabetes,  `r table.agegr.all[6, 3]` for Type 2 diabetes, 
`r table.agegr.all[8, 3]` for ischemic heart disease,
`r table.agegr.all[9, 3]` for other heart disease, 
`r table.agegr.all[10, 3]` for chronic lower respiratory tract disease, 
`r table.agegr.all[11, 3]` for chronic kidney disease, 
`r table.agegr.all[12, 3]` for neurological disease, 
`r table.agegr.all[13, 3]` for chronic liver disease and 
`r table.agegr.all[14, 3]` for immune deficiency or suppression.

`r freqs.all[nrow(freqs.all), 2]` of the severe cases and `r freqs.all[nrow(freqs.all), 1]` of the controls had at least one of these listed conditions.  In those aged under 40 years the frequencies of any listed condition were `r gsub("\\\\", "", table.agegr[4, 2])` in cases and `r gsub("\\\\", "", table.agegr[4, 1])` in controls.  The odds ratio for severe disease associated with encashment of at least one prescription in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`; these associations remained strong and significant after adjusting for the listed conditions.  Most cases without listed conditions had either prescriptions or hospitalisations, and highly significant associations with severe disease were seen across many hospital diagnoses and drug categories.

A predictive  model based on age and sex alone had an expected information for discrimination of  `r round(cases.Lambda.agesex, 2)` bits.   In the model conditional on age and sex further demographic variables provided another `r table.densities[1, 5]` bits,  and this increased to `r table.densities[2, 5]` bits  when listed conditions were included, and to `r table.densities[3, 5]` bits when indicator variables for hospital diagnoses and prescription encashments were also included.

_Conclusions_ --Along with older age and male sex, severe COVID-19 disease is strongly associated with past medical history; most younger persons developing severe disease have an underlying medical condition. The comorbidities associated with increased risk are not limited to the lists of risk conditions designated by public health agencies.  A risk classifier that uses all the information available in medical and prescribing records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the SARS-CoV-2 epidemic is over.

@@

The eight listed conditions were scored based on diagnostic codes in any hospital discharge record during the last five years, or encashed prescription of a drug for which the only indications are in that group of diagnostic codes. The R script included as supplementary material contains the derivations of these variables from ICD-10 codes and BNF drug codes.  Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other/Unknown. Recent cases of diabetes since the last update of the register were supplemented with  a small number new cases identified through discharge codes and drug codes.  These new cases were added into the diabetes type Other/Unknown category.

@@

To investigate this further, we partitioned the "Any prescription" variable into indicator variables for each chapter of the British National Formulary, in which drugs are grouped by broad indication, and restricted the analysis to those without one of the listed conditions.  Table \ref{tab:drugs} shows these associations.  In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease.  In a multivariate analysis of all chapters, most of these associations were weaker.  The BNF chapters with the strongest independent associations with severe disease were chapters 1 (gastrointestinal) and 2 (cardiovascular)).  Other chapters associated with severe disease were chapters 4 (central nervous system), 9 (nutrition and blood) and 14+ (other, mostly dressings and appliances).

@@

Media reports of apparently healthy young people succumbing to severe COVID-19 have disseminated the message that all are at risk of disease whatever their age or health status.  However we found that almost half of cases under 40 years had at least one of the listed conditions and among those who did not have one of these conditions, the proportions who had at least one prior hospitalisation or dispensed prescription were much higher in cases than in controls.  In all age groups, very few of the fatal cases had not had either a hospital admission in the past five years or a dispensed prescription in the past year.

@@

rownames(table.agegr) <- replace.names(rownames(table.agegr))
#table.agegr[1, c(1, 3, 5, 7)] <- " "
knitr::kable(table.agegr[-7, ], # omit non-diabetic category
             escape=FALSE, 
             booktabs=TRUE,
             label="agegr",
             align=rep("c", 8), 
             caption="Frequencies of risk factors in cases and controls, by age group") %>%
    column_spec(column=1, width="4cm") %>%
    column_spec(column=2:9, width="2cm") %>%
    add_header_above(c(" "=1, "0-39 years"=2, "40-59 years"=2,
                       "60-74 years"=2, "75+ years"=2)) %>%
 #   kable_styling(latex_options="scale_down") %>%
    landscape() %>%
    kable_styling(latex_options="hold_position")

>>>>

_Background_--   The objectives of this study were to identify risk factors for severe  COVID-19 and to lay the basis for risk stratification based on demographic data and health records.

_Methods_ -- The design was a matched case-control study. Severe cases were all those with a positive nucleic acid test for SARS-CoV-2 in the national database who had entered a critical care unit or died within 28 days of the first positive test.  Ten controls per case matched for sex, age and primary care practice were selected from the population register.  All diagnostic codes from the past five years of hospitalisation records and all drug codes from prescriptions dispensed during the past nine months were extracted.  Rate ratios for severe COVID-19 were estimated by conditional logistic regression.

_Findings_ -- In a logistic regression using the age-sex distribution of the national population, the odds ratios were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.  In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]`.  Univariate rate ratios (95% CIs) for conditions listed by public health agencies as conferring high risk were `r table.agegr.all[5, 3]` for Type 1 diabetes, `r table.agegr.all[6, 3]` for Type 2 diabetes, 
`r table.agegr.all[8, 3]` for ischemic heart disease,
`r table.agegr.all[9, 3]` for other heart disease, 
`r table.agegr.all[10, 3]` for chronic lower respiratory tract disease, 
`r table.agegr.all[11, 3]` for chronic kidney disease, 
`r table.agegr.all[12, 3]` for neurological disease, 
`r table.agegr.all[13, 3]` for chronic liver disease and 
`r table.agegr.all[14, 3]` for immune deficiency or suppression.

`r gsub("([0-9]+ \\()(.*)(\\))", "\\2", freqs.all[nrow(freqs.all), 2])` of cases and `r gsub("([0-9]+ \\()(.*)(\\))", "\\2", freqs.all[nrow(freqs.all), 1])` of controls had at least one listed condition (`r gsub("([0-9]+ \\()(.*)(\\))", "\\2", table.agegr[4, 2])` of cases and `r gsub("([0-9]+ \\()(.*)(\\))", "\\2", table.agegr[4, 1])` of controls under age 40).  Severe disease was associated with encashment of at least one prescription in the past nine months and with at least one hospital admission in the past five years [rate ratios `r table.agegr.all[2, 3]`] and `r table.agegr.all[3, 3]` respectively] even after adjusting for the listed conditions.  In those without listed conditions significant associations with severe disease were seen across many hospital diagnoses and drug categories. Age and sex provided `r round(cases.Lambda.agesex, 2)` bits of information for discrimination.  A model based on demographic variables, listed conditions, hospital diagnoses and prescriptions provided an additional `r table.densities[3, 5]` bits (C-statistic `r table.densities[3, 3]`).

_Conclusions_ -- Along with older age and male sex, severe COVID-19 is strongly associated with past medical history across all age groups. Many comorbidities beyond the risk conditions designated by public health agencies contribute to this.  A risk classifier that uses all the information available in health records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the epidemic is over.

@@

The eight listed conditions were scored based on diagnostic codes in any hospital discharge record during the last five years, or encashed prescription of a drug for which the only indications are in that group of diagnostic codes. The R script included as supplementary material contains the derivations of these variables from ICD-10 codes and BNF drug codes.  Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other/Unknown. Cases of diabetes diagnosed since the last update of the register were identified through discharge codes and drug codes, and assigned to the diabetes type Other/Unknown category.

@@

To investigate this further, we partitioned the "Any prescription" variable into indicator variables for each chapter of the British National Formulary, in which drugs are grouped by broad indication, and restricted the analysis to those without one of the listed conditions.  Table \ref{tab:drugs} shows these associations.  In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease.  In a multivariate analysis of all chapters, most of these associations were weaker.  The BNF chapters with the strongest independent associations with severe disease were chapters 1 (gastrointestinal) and 2 (cardiovascular).  Other chapters associated with severe disease were chapters 4 (central nervous system), 9 (nutrition and blood) and 14+ (other, mostly dressings and appliances).

@@

Media reports of apparently healthy young people succumbing to severe COVID-19 have disseminated the message that all are at risk of disease whatever their age or health status.  However we found that half of cases under 40 years had at least one of the listed conditions and among those who did not have one of these conditions, the proportions who had at least one prior hospitalisation or dispensed prescription were much higher in cases than in controls.  In all age groups, very few of the fatal cases had not had either a hospital admission in the past five years or a dispensed prescription in the past year.

@@

rownames(table.agegr) <- replace.names(rownames(table.agegr))
knitr::kable(table.agegr[-7, ], # omit non-diabetic category
             escape=FALSE, 
             booktabs=TRUE,
             label="agegr",
             align=rep("c", 8), 
             caption="Frequencies of risk factors in cases and controls, by age group") %>%
    column_spec(column=1, width="4cm") %>%
    column_spec(column=2:9, width="2cm") %>%
    add_header_above(c(" "=1, "0-39 years"=2, "40-59 years"=2,
                       "60-74 years"=2, "75+ years"=2)) %>%
    landscape() %>%
    kable_styling(latex_options="hold_position")
